% Abstract — pivoted framing (the DSL kernel *evaluation* problem).
% Canonical framing: ../PIVOT_FRAMING.md. Pre-pivot gap-study drafts are in git history.
Modern GPU domain-specific languages (DSLs) such as Triton and TileLang deliver competitive performance relative to vendor libraries for GPU kernel programming,
making them a popular choice for human developers and an increasingly common target for automated kernel code generation agents.
Yet developer-written or agent-generated DSL kernels can be \emph{functionally correct} but offer no dependable
way to know whether they are \emph{performant}.
We show that a kernel can pass the prevailing evaluation benchmarks and still run
hundreds of times slower than the vendor-library baseline: across 22 kernels in five
 categories, naive-but-idiomatic DSL kernels that pass correctness
gating range from competitive to more than two orders of magnitude slower than
cuBLAS/cuDNN.

The benchmarks practitioners rely on do not surface this.
Prevailing benchmarks (e.g., KernelBench and TritonBench)
gate primarily on correctness and therefore
admit performance-poor kernels, while under-covering DSLs, data types, shapes, and
hardware and only partially preventing reward hacking.
A benchmark comprehensive enough to close this gap is combinatorially infeasible.

As a pragmatic alternative, we (i)~characterize how large and structured the hidden
performance gap is and why it arises, separating user-space \emph{authoring} causes
from \emph{code-generation} and
\emph{library-maturity} causes; and (ii)~distill the result into
lightweight, partly baseline-independent evaluation heuristics and a small set of
recurring optimization patterns that guide DSL kernel development when no ground-truth
benchmark exists.
We find that authoring idiom dominates the gap: a single reduction-primitive change 
recovers the normalization and reduction family to library parity or better 
(with one hardware-specific exception we diagnose), while a residual structural gap 
persists on convolution. We further show that a roofline-anchored comparability check 
flags poor kernels that correctness-only benchmarks pass.
Our kernel suite, profiling infrastructure, and analysis are publicly available at
\url{https://anonymous.4open.science/r/GPUKernelPerformance-6132/}.
