\subsection{Implications for DSL Developers}
\label{sec:disc:dsl}

% Our findings suggest three concrete improvements to Triton's compiler backend.

% \paragraph{Vectorization of strided spatial accesses.}
% The vectorization failure identified in RC1 (\cref{sec:analysis:vec})
% is amenable to a targeted fix:
% Triton's alias analysis pass could be extended to reason about
% \emph{channel-last} (NHWC) access patterns,
% emitting \texttt{LDG.128} unconditionally when the channel dimension
% is the contiguous dimension and the channel count is a multiple of 8 (for FP16).
% This is a compiler change, not a user-visible API change.

% \paragraph{Problem-shape-aware auto-tuning.}
% The mismatch between Triton's static configuration lists and the
% convolution search space (RC2) calls for a more adaptive search strategy.
% Ansor~\cite{zheng2020ansor} demonstrates that hierarchical program sketches
% can discover high-performance configurations for irregular operator shapes (including convolution)
% without manual specification, outperforming template-guided TVM auto-tuning
% by up to $3.8\times$ on representative workloads in their published evaluation.
% A similar adaptive search mechanism in Triton's auto-tuner ---
% replacing the static configuration list with a shape-aware search strategy analogous to Ansor's sketch-based approach adapted to Triton's IR ---
% is a concrete engineering path to closing RC2 (\cref{sec:analysis:autotune}).

% \paragraph{Scheduling primitives for Winograd.}
% Exposing the Winograd data transformation as a first-class TileLang tile operator
% would enable the DSL's scheduling infrastructure to select among GEMM-lowering,
% Winograd, and FFT paths at tuning time,
% analogously to cuDNN's runtime algorithm selector.
% This is a larger engineering effort but would close the algorithmic gap (RC4)
% for $3 \times 3$ convolution workloads.

Our results point to three practical directions for improving current GPU DSL systems.

\textbf{Improve vectorization for convolution accesses (requires compiler change).}
RC1 (\cref{sec:analysis}) shows that convolution performance is limited in part by the failure to emit wide \texttt{LDG.128} loads for strided spatial accesses. A direct compiler remedy is to extend alias and layout analysis to recognize the convolution access pattern and emit vectorized loads when the contiguous dimension is aligned (e.g., multiples of 8 for FP16). This change targets code generation rather than the programming model, and therefore does not require changes to the user-facing API.

\textbf{Make auto-tuning shape-aware (partly user-space fixable today; broader search requires compiler change).}
RC2 (\cref{sec:analysis}) indicates that static tuning configurations transfer poorly from GEMM-like workloads to convolution. The RC2a portion is fixable in user space today: manually expanding the configuration list (e.g., \texttt{GROUP\_SIZE\_M}/\texttt{num\_warps}/\texttt{num\_stages}) recovers most of the GEMM gap, as shown in \cref{sec:mitigation}. Closing the residual RC2b gap calls for auto-tuning strategies that are more adaptive to operator shape and memory-access structure. Prior work such as Ansor~\cite{zheng2020ansor} shows that sketch-based search can discover effective schedules for irregular operators, including convolution, without relying on manually enumerated templates. A similar shape-aware search mechanism for Triton or TileLang would be a concrete step toward reducing this gap.
Our own sweep illustrates both the hazard and its cost: tuning at a shape smaller than deployment can select a configuration \emph{slower than the untuned default} (\cref{sec:threats}), and tuning at the deployment shape---the obvious remedy---is expensive, because each candidate must be timed on the full-size problem.
Re-tuning a single slow kernel (attention) at its deployment shape took $211$\,s versus under a second at the small shape---a ${\sim}1000\times$ difference that compounds across kernels, dtypes, and shapes.
Shape-representative tuning is therefore not free; this reinforces both that an exhaustively-tuned comprehensive benchmark is infeasible (\cref{sec:evaluation}) and that the lightweight comparability-plus-roofline heuristics of \cref{sec:mitigation} are the more practical screen.

\textbf{Support algorithmic alternatives for convolution (requires compiler change).}
RC4 highlights a limitation that cannot be addressed through local schedule tuning alone: current DSL implementations do not expose algorithmic choices comparable to those available in cuDNN. In particular, support for Winograd-style transformations would allow the compiler or runtime to choose among direct convolution, GEMM lowering, and Winograd-based execution depending on filter shape and problem size. Although this requires a larger system change than RC1 or RC2, and the wall-clock benefit attributable specifically to Winograd is small ($\approx$2--3\%), it contributes to closing the remaining gap on common $3\times3$ convolutions.


% \subsection{Implications for Practitioners}
% \label{sec:disc:practitioners}

% \paragraph{Don't assume library parity.}
% Our results confirm that switching from PyTorch (which delegates to cuDNN)
% to a hand-written Triton convolution kernel does not automatically improve performance.
% For element-wise and normalization kernels, the switch can be productivity-positive with no throughput loss (Triton achieves 95--103\% of PyTorch throughput on LayerNorm and element-wise kernels; RMSNorm exceeds 1099\% due to kernel fusion and is not representative of the category);
% for GEMM, modest throughput costs (31--58\% of cuBLAS throughput on large shapes) may be acceptable when the custom operation cannot be expressed as an existing library call;
% for convolution, a substantial regression (35\% for Triton, 9\% for TileLang) should be expected unless mitigations are applied (\cref{sec:mitigation}).
% Teams adopting Triton for conv-heavy workloads (ResNets, CNNs, 3D segmentation models)
% should profile against a cuDNN baseline before shipping.

% \paragraph{Use the root-cause taxonomy as a checklist.}
% The targeted mitigations in \cref{sec:mitigation} can be applied as a checklist
% when a Triton or TileLang convolution kernel under-performs:
% first check vectorized-load fraction in Nsight Compute (RC1),
% then sweep a broader tile configuration space (RC2),
% then check register spill (RC3).
% For $3 \times 3$ filters, evaluate whether Winograd overhead is justified (RC4).

The results also have immediate consequences for users who adopt DSLs to replace library-backed kernels.

\textbf{Performance depends strongly on operator class.}
DSL kernels should not be assumed to match library performance by default. In our study, Triton is near parity on element-wise and normalization workloads, and can therefore offer a favorable trade-off between programmability and performance in these cases. For GEMM, the observed cost relative to cuBLAS may be acceptable when customization or fusion is required. Convolution is different: both Triton and TileLang remain substantially behind cuDNN in the evaluated settings. For convolution-heavy models, DSL kernels should therefore be treated as an optimization target that must be validated against a library baseline, not as a drop-in replacement.

\textbf{The root-cause analysis can guide debugging.}
The mitigation results suggest a practical workflow for diagnosing underperforming DSL kernels. For TileLang reductions and normalization, developers should first replace serial \texttt{T.serial} accumulation with \texttt{T.reduce} (RC0a), since the dominant cost is exposed memory latency rather than thread synchronization; this is also where register pressure and spill should be inspected (RC3), which on our hardware appears in TileLang normalization rather than in convolution. For convolution, developers should check whether memory accesses are vectorized into wide \texttt{LDG.128} loads (RC1), then evaluate whether the tuning configuration space is too narrow for the target shape, distinguishing a config that is merely absent from the default grid (RC2a, fixable by expanding the search) from one that needs a broader search strategy (RC2b). For $3\times3$ filters, Winograd selection (RC4) is worth considering, but only as a small ($\approx$2--3\%) contributor. This sequence provides a compact checklist for performance debugging in Triton and TileLang.


\textbf{Benchmarks should gate on performance, not just correctness.}
The passes-but-slow result (\cref{sec:eval:passesslow}) carries a concrete recommendation for benchmark designers: a correctness gate is necessary but not sufficient, and a kernel benchmark that reports speedup without gating on it certifies the wrong property. A lightweight, baseline-independent roofline check (\cref{sec:mitig:heuristics}) is a practical acceptance criterion that needs no curated fast reference, and would have flagged every passes-but-slow kernel in our study.


\subsection{Limitations}
\label{sec:disc:limitations}

% Our study is bounded by the following factors.

% \paragraph{Kernel scope.}
% We focus on inference-time forward-pass kernels.
% Backward-pass gradient kernels introduce additional complexity
% (larger temporary buffers, more accumulation loops)
% and are left for future work.

% \paragraph{Hardware scope.}
% We test primarily on an NVIDIA A100-SXM4-40GB (sm_80), with cross-arch replay on A100-PCIE-40GB and H100-80GB-HBM3.
% AMD ROCm and Intel XPU are excluded;
% TileLang's AMD support~\cite{wang2025tilelang} makes AMD an important target for future work.

% \paragraph{DSL versions.}
% Triton and TileLang are under active development.
% Compiler improvements released after our study's software snapshot
% (see \cref{sec:meth:setup}) may partially address the gaps we identify.
% Our benchmark suite is designed to make re-evaluation straightforward.

Our conclusions should be interpreted within the scope of the study.

\textbf{Operator scope.}
We evaluate forward-pass kernels only. Backward kernels introduce different computation and memory patterns, including larger temporary state and more complex accumulation behavior, and may expose additional bottlenecks not captured here.

\textbf{Hardware scope.}
Our primary measurements span two architectures, the NVIDIA A100-SXM4-40GB (Ampere, sm\_80) and the NVIDIA GH200 (Grace Hopper, sm\_90), with cross-architecture replay on the A100-PCIE-40GB (same sm\_80) and H100-80GB-HBM3 confirming that the gaps generalize across form factors and GPU families. The results nonetheless characterize NVIDIA datacenter platforms rather than all accelerators; in particular, we do not evaluate ROCm or Intel XPU backends, although these are important targets for future work.

\textbf{Software snapshot.}
Both Triton and TileLang are evolving systems. The measurements in this paper reflect the software versions listed in \cref{sec:meth:setup}; later compiler or runtime changes may reduce some of the gaps we report. Our released benchmark suite is intended to support such re-evaluation.