\subsection{Kernel Suite}
\label{sec:meth:kernels}

% We collect a kernel suite spanning five operator categories
% that cover the dominant computational patterns in modern deep learning.
% In total, our suite comprises 22 kernels:
% 3 \textbf{GEMM} (dense matmul, batched matmul, fused linear+activation),
% 1 \textbf{Attention} (scaled dot-product / FlashAttention variants),
% 1 \textbf{Convolution} (Conv2d, $1\times1$--$7\times7$ filters, depthwise, strided),
% 2 \textbf{Normalization} (LayerNorm, RMSNorm),
% and 15 \textbf{element-wise and reduction} operators
% (add, mul, relu, leaky\_relu, softmax, log\_softmax, logsumexp, swiglu,
%  argmax, max reduction, mean reduction, cross-entropy, matrix transpose, index select, embedding).

% Our 22 kernels are sourced from TritonBench~\cite{li2025tritonbench},
% the TileLang example repository~\cite{wang2025tilelang},
% and custom implementations for configurations absent from both suites.
% We exclude kernels with sparse inputs or runtime-dependent output shapes,
% as these lack meaningful library baselines.

Our benchmark suite covers five operator categories that capture the dominant computation patterns in modern deep learning: matrix multiplication, attention, convolution, normalization, and element-wise or reduction operators. In total, the suite contains 22 kernels: three \textbf{GEMM} workloads (i.e., dense matrix multiplication, batched matrix multiplication, and fused linear plus activation), one \textbf{attention} workload (i.e., scaled dot-product attention / FlashAttention variants), one \textbf{convolution} workload (i.e., Conv2d with $1\times1$ to $7\times7$ filters, including depthwise and strided cases), two \textbf{normalization} workloads (i.e., LayerNorm and RMSNorm), and fifteen \textbf{element-wise or reduction} workloads including add, mul, relu, leaky\_relu, softmax, log\_softmax, logsumexp, swiglu, argmax, max reduction, mean reduction, cross-entropy, matrix transpose, index select, and embedding.

We derive these workloads from TritonBench~\cite{li2025tritonbench}, the TileLang example repository~\cite{wang2025tilelang}, and our own implementations for configurations not covered by either source. We exclude operators with sparse inputs or runtime-dependent output shapes, since they do not admit stable or meaningful library baselines.

% REVISION-NOTE-START id=4134A-W6 severity=Major
\textcolor{red}{\textbf{[REVISION \#4134A-W6 + \#GAP-7, Major]}
Concern (\textit{Reviewer A}): \cref{sec:meth:kernels} claims ``Conv2d with filter sizes from $1\times1$ to $7\times7$, including depthwise and strided variants'' but \cref{tab:conv} reports only the $3\times3$ stride-1 configuration.
Evidence: \texttt{../experiments/results/NVIDIA\_RTX\_4000\_Ada\_Generation/conv\_filters.csv}.
Fix: Expand \cref{tab:conv} to include $1\times1$, $5\times5$, $7\times7$, depthwise, and strided rows drawn from \texttt{conv\_filters.csv}, or soften the claim here to match the table coverage.}
% REVISION-NOTE-END id=4134A-W6

% REVISION-NOTE-START id=4134A-W9 severity=Medium
\textcolor{red}{\textbf{[REVISION \#4134A-W9 + \#4134B-Q1, Medium]}
Concern (\textit{Reviewers A and B}): Reviewers ask for explicit selection criteria and per-kernel provenance for the 22-kernel suite.
Evidence: \texttt{../REBUTTAL.md} \S1 A\#9.
Fix: Add a 4-bullet selection-criteria block (forward-pass only; stable library baseline; five-category coverage; canonical ops --- no sparse or dynamic shapes) and a provenance table mapping each kernel to its source (Triton kernels from TritonBench, \texttt{layer\_norm} from TorchInductor, TileLang re-implementations).}
% REVISION-NOTE-END id=4134A-W9


\subsection{Baseline Construction}
\label{sec:meth:baselines}

% Library baselines use PyTorch dispatch to cuBLAS (\texttt{torch.matmul} for GEMM),
% \texttt{nn.Conv2d} in NHWC layout with \texttt{allow\_tf32=False} for convolution,
% \texttt{F.layer\_norm}/\texttt{F.rms\_norm} for normalization,
% and PyTorch eager execution for element-wise operators.
% Attention baselines use scaled dot-product attention with the FlashAttention backend
% (\texttt{enable\_flash\_sdp(True)}); results are presented separately because
% the Triton and TileLang implementations differ algorithmically from PyTorch's baseline.

For each workload, we compare DSL implementations against the corresponding vendor-library path exposed through PyTorch. GEMM baselines use cuBLAS via \texttt{torch.matmul}. Convolution baselines use \texttt{nn.Conv2d} in NHWC layout with \texttt{allow\_tf32=False}. Normalization baselines use \texttt{F.layer\_norm} and \texttt{F.rms\_norm}. Element-wise and reduction baselines use standard PyTorch eager execution.

% REVISION-NOTE-START id=Internal-N1 severity=Major decision=true
\textcolor{red}{\textbf{[REVISION \#Internal-N1 + \#GAP-6, Major, AUTHOR DECISION]}
Concern (\textit{Internal}): Paper claims \texttt{cudnn.benchmark=False}, \texttt{allow\_tf32=False}, NHWC layout; \texttt{ViperBench/benchmark.py} sets none of them.
Evidence: \texttt{../ViperBench/benchmark.py}, \texttt{../ViperBench/conv2d/triton\_impl.py}.
Option A: Update \texttt{benchmark.py} to set the three flags + \texttt{channels\_last} and re-run \cref{tab:conv} / \cref{tab:mitig:conv}.
Option B: Soften this sentence to ``PyTorch baselines use cuBLAS, default-layout Conv2d under default cuDNN flags.''
Decision needed before camera-ready.}
% REVISION-NOTE-END id=Internal-N1

For attention, we use PyTorch scaled dot-product attention with the FlashAttention backend enabled through \texttt{enable\_flash\_sdp(True)}. We report these results separately because the Triton and TileLang implementations are not algorithmically identical to the PyTorch baseline.


\subsection{Profiling Setup}
\label{sec:meth:profiling}

% \paragraph{End-to-end throughput.}
% For each (kernel, configuration) pair we measure wall-clock kernel time
% using CUDA events (\texttt{cudaEvent\allowbreak Record} / \texttt{cuda\allowbreak Event\allowbreak Elapsed\allowbreak Time}),
% averaged over 100 iterations.
% Throughput measurements warm up with 10 un-timed iterations before 100 timed iterations.
% We report GPU-side time only, excluding host-side dispatch overhead.
% All kernels are run in isolation (no concurrent workloads)
% to minimize measurement noise from GPU frequency scaling.

% \paragraph{Hardware performance counters.}
% For root-cause analysis (RQ2) we collect hardware counter profiles
% using NVIDIA Nsight Compute~\cite{nsightcompute2024}.
% We collect three counter groups via Nsight Compute: \emph{memory access} (global load/store efficiency, L1/L2 hit rates, coalescing, vectorized-load fraction), \emph{compute utilization} (SM occupancy, warp stall cycles, Tensor Core utilization), and \emph{instruction mix} (\texttt{cp.async} issue count, register spill).
% Each kernel is profiled in a single iteration; stability is verified within 2\% over five repeats.

We measure both end-to-end latency and hardware performance counters. Latency is used for all comparisons against library baselines, while hardware counters are used to investigate the causes of observed performance gaps.

\paragraph{End-to-end throughput.}
For each workload and input configuration, we measure GPU execution time using CUDA events (\texttt{cudaEventRecord} and \texttt{cudaEventElapsedTime}). Each measurement uses 10 warm-up iterations followed by 100 timed iterations, and we report the average over the timed runs. All results reflect GPU-side execution only and exclude host-side dispatch overhead. To reduce measurement noise, we run each workload in isolation without concurrent GPU activity.

% REVISION-NOTE-START id=GAP-5 severity=Major
\textcolor{red}{\textbf{[REVISION \#GAP-5, Major]}
Concern (\textit{Internal}): \texttt{ViperBench/benchmark.py:41-48} uses \texttt{time.perf\_counter()} and reports the \textbf{median} of 100 timed runs, not CUDA events and not the average. The current description does not match the code that produced \texttt{profile.csv}.
Evidence: \texttt{../ViperBench/benchmark.py:41-48}.
Fix: Replace the paragraph with: ``we measure host-synchronized GPU execution time using \texttt{time.perf\_counter()} bracketed by \texttt{torch.cuda.synchronize()}, after 10 warm-up iterations followed by 100 timed iterations; we report the median of the timed runs.'' Keep CUDA-event wording only if Option A in \#Internal-N1 is chosen and \texttt{benchmark.py} is changed accordingly.}
% REVISION-NOTE-END id=GAP-5

\paragraph{Hardware performance counters.}
To support root-cause analysis, we collect hardware counter profiles with NVIDIA Nsight Compute~\cite{nsightcompute2024}. We group the collected counters into three classes: \emph{memory access} (e.g., global load/store efficiency, L1/L2 hit rates, coalescing behavior, and vectorized-load fraction), \emph{compute utilization} (e.g., SM occupancy, warp stall cycles, and Tensor Core utilization), and \emph{instruction behavior} (e.g., \texttt{cp.async} issue count and register spill). Each profile is collected from a single execution, and we verify stability within 2\% across five repeated runs.

% REVISION-NOTE-START id=4134A-W3 severity=Major
\textcolor{red}{\textbf{[REVISION \#4134A-W3, Major]}
Concern (\textit{Reviewer A}): Reviewers explicitly asked for load-vec\%, register spill, warp-stall, and L2/DRAM counters; this paragraph only lists counter \emph{categories}, not the specific counters that were collected.
Evidence: \texttt{../experiments/results/NVIDIA\_RTX\_4000\_Ada\_Generation/ncu\_summary.csv}.
Fix: List the seven concrete counters used in the RQ2 analysis: \texttt{gld\_efficiency}, \texttt{ldg\_per\_request}, \texttt{n\_regs}, \texttt{n\_spills}, \texttt{warp\_stall\_long\_scoreboard}, \texttt{warp\_stall\_barrier}, and \texttt{lts\_\_t\_sector\_hit\_rate}.}
% REVISION-NOTE-END id=4134A-W3

% REVISION-NOTE-START id=4134A-W11 severity=Major
\textcolor{red}{\textbf{[REVISION \#4134A-W11 + \#GAP-5, Major]}
Concern (\textit{Reviewer A}): Correctness validation is under-described --- no per-dtype tolerances, no edge cases, and no FP32 root cause are stated.
Evidence: \texttt{../ViperBench/test\_utils.py}, \texttt{../experiments/results/NVIDIA\_RTX\_4000\_Ada\_Generation/correctness\_edge.csv}.
Fix: Add a Correctness subsection stating per-dtype tolerances (\texttt{fp32: 1e-5}, \texttt{fp16: 1e-3}, \texttt{bf16: 1e-2}) with \texttt{loose\_tol=True} ($\times2$) for reductions; an edge-case suite (NaN/Inf/denormal/all-equal) over 22 kernels with 0 crashes; mitigation revalidation (5/5 pass); and note that FP32 GEMM is excluded since it is root-caused as silent TF32 lowering (forward ref to \cref{sec:analysis:jit}).}
% REVISION-NOTE-END id=4134A-W11


\subsection{Metrics}
\label{sec:meth:metrics}

% The primary metric is \textbf{library efficiency}
% $E_\text{lib} = t_\text{lib} / t_\text{DSL} \times 100\%$,
% where $t_\text{lib}$ is the cuBLAS/cuDNN baseline latency and $t_\text{DSL}$ is the DSL kernel latency
% on identical hardware and shape; 100\% indicates parity, below 100\% means the DSL is slower.
% Secondary metrics---throughput ($T = 2 \cdot \text{FLOPs}/t$), hardware efficiency
% ($E_\text{hw} = T/T_\text{peak}$), and memory bandwidth utilization
% ($B = \text{bytes}/({t \cdot B_\text{peak}})$)---are reported for root-cause analysis (RQ2).

Our primary metric is \textbf{library efficiency},
\[
E_{\mathrm{lib}} = \frac{t_{\mathrm{lib}}}{t_{\mathrm{DSL}}} \times 100\%,
\]
where $t_{\mathrm{lib}}$ denotes the latency of the cuBLAS or cuDNN baseline and $t_{\mathrm{DSL}}$ denotes the latency of the corresponding DSL implementation under the same hardware and input configuration. A value of 100\% indicates parity with the library baseline; lower values indicate that the DSL implementation is slower.

Secondary metrics: throughput ($T = 2 \cdot \text{FLOPs}/t$), hardware efficiency
($E_\text{hw} = T/T_\text{peak}$), and memory bandwidth utilization
($B = \text{bytes}/({t \cdot B_\text{peak}})$)---are used qualitatively in root-cause analysis (RQ2)
to interpret counter measurements; they are not tabulated separately.

\subsection{Experimental Setup}
\label{sec:meth:setup}

% \paragraph{Hardware.}
% All experiments run on an NVIDIA RTX 4000 Ada Generation (Ada Lovelace, \texttt{sm\_89}),
% with 20~GB GDDR6 memory (360~GB/s peak bandwidth) and a 48~MB L2 cache.
% NVIDIA Driver 595, CUDA Toolkit 12.6.

% \paragraph{Software.}
% PyTorch 2.8.0+cu126, Triton 3.4.0, TileLang 0.1.6.post1, cuDNN 9.1.0.2, Python 3.9.25,
% Nsight Compute 2024.3.2.0; all versions are pinned in a reproducibility manifest
% distributed with the benchmark suite.
% We set random seeds, disable cuDNN benchmark mode (\texttt{torch.backends.cudnn.benchmark=False}),
% and pre-warm the Triton auto-tuner before measurement.

\paragraph{Hardware.}
All experiments are conducted on an NVIDIA RTX 4000 Ada Generation GPU (Ada Lovelace, \texttt{sm\_89}) with 20\,GB GDDR6 memory, 360\,GB/s peak memory bandwidth, and a 48\,MB L2 cache. The system uses NVIDIA Driver 595 and CUDA Toolkit 12.6.

% REVISION-NOTE-START id=4134A-W13 severity=Minor
\textcolor{red}{\textbf{[REVISION \#4134A-W13, Minor]}
Concern (\textit{Reviewer A}): Elsewhere (\cref{sec:analysis:regpressure} RC3 section), the paper cites A100 register characteristics in this Ada-only study; the value is correct (Ada and A100 both have 65{,}536 32-bit registers per SM) but the citation should be Ada-localized.
Evidence: \texttt{../paper-latex-project/GPU Kernel Performance Analysis Report.md}.
Fix: When the RC3 section cites A100 register file size, change to Ada (\texttt{sm\_89}) or add ``Ada and A100 share the 65{,}536-register-per-SM budget.'' This note flags the architecture spec here; the actual change happens in \cref{sec:analysis:regpressure}.}
% REVISION-NOTE-END id=4134A-W13

\paragraph{Software.}
Our software stack consists of PyTorch 2.8.0+cu126, Triton 3.4.0, TileLang 0.1.6.post1, cuDNN 9.1.0.2, Python 3.9.25, and Nsight Compute 2024.3.2.0. We pin all versions in a reproducibility manifest distributed with the benchmark suite. For all experiments, we fix random seeds, disable cuDNN benchmark mode (\texttt{torch.backends.cudnn.benchmark=False}), and pre-warm the Triton auto-tuner before measurement.

% REVISION-NOTE-START id=Internal-N1 severity=Major decision=true
\textcolor{red}{\textbf{[REVISION \#Internal-N1 + \#GAP-6, Major, AUTHOR DECISION]}
Concern (\textit{Internal}): Paper claims \texttt{cudnn.benchmark=False} is set, but \texttt{ViperBench/benchmark.py} sets nothing. This is the same condition as the \cref{sec:meth:baselines} note --- both must resolve to the same option.
Evidence: \texttt{../ViperBench/benchmark.py}.
Option A: Update \texttt{benchmark.py} to set \texttt{cudnn.benchmark=False}, \texttt{allow\_tf32=False}, and \texttt{channels\_last} layout, then re-run \cref{tab:conv} / \cref{tab:mitig:conv}.
Option B: Soften this sentence to remove the claim that \texttt{cudnn.benchmark=False} and seed-fixing are set, and instead state that experiments run under default cuDNN flags.
Decision needed before camera-ready; must match the \cref{sec:meth:baselines} resolution.}
% REVISION-NOTE-END id=Internal-N1
